Glucose ingestion before and after resistance training does not augment ribosome biogenesis in healthy moderately trained young adults

### 2023-02-08 ribose-paper.RMD
updated intro

Regular resistance exercise training (RT) increases muscle mass and muscle strength, due to the adaptations stimulated by  mechanical strain. Unfortunately, responses to RT are not uniform across the population. Studies have shown RT-induced muscle growth to vary widely between individuals, with as much as 10-15% showing considerably impaired growth (ALVAREZ18, MANN14, THALACKER13). This may be explained by genetics, epigenetics, and/or an unfavourable internal physiological milieu (THALACKER13, HAMMARSTROM20, FIGUEIREDO19?). While tweaking training modalties and -protocols induces different responses in different populations (HAMMARSTROM20, 19?), little is known regarding optimisation of RT to individual phenotypes. Generally, meta-analyses favour a moderate volume of RT over low volume RT regarding muscle growth and -strength (KRIEGER09, RHEA03, SCHOENFELD17), substantiated by recent RCTs (HAMMARSTROM20..). Although Hammarström and colleagues (20) found that a moderate volume led to more prominent gains in muscle mass and -strength, approximately 50% of the participants did not experience true benefits of moderate volume compared to low, with a blunted muscle growth. Due to the variations observed in muscle growth responses despite standardising RT protocols (HAMMARSTROM20; SCHOENFELD17; STEC16?), it is apparent that increasing RT volume alone does not convert low-responders to high-responders. Therefore, other means than RT *per se* seems necessary to circumvent this discrepancy.

Indeed, ingesting nutrients such as protein- and creatine-supplements effectively optimise RT (CERMAK12, LANHERS15, LANHERS17). However, it remains unknown if other nutritional adjuvants such as glucose can increase the efficacy of RT. This is surprising, as glucose is the preferred energy substrate of the contracting skeletal muscle during strenuous exercise and a major supplier of energy to cells via ATP synthesis (TANAKA18). Energy availability is important to regulate central processes of muscle growth, specifically ribosome biogenesis, a key factor for increased muscle protein synthesis (MPS) (TANAKA18, FIGUEIREDO19, VONWALDEN19?). Interestingly, variations in RT-induced responses have been linked to ribosome biogenesis, where participants responding poorly to RT also showed a blunted ribosome biogenesis (HAMMARSTROM20, STEC16). A blunted ribosome biogenesis negates responses to RT by failure to produce novel ribosomes, thus an inability to increase the skeletal muscle cell’s translational capacity (VONWALDEN19, FIGUEIREDO19). Therefore, investigating the effects of combined RT and glucose ingestion may provide valuable insight into the regulation of ribosome biogenesis.

Ribosome biogenesis is suggested to be regulated by the mammalian target of rapamycin complex 1 (mTORC1) network (KUSNADI15, TANAKA18). Stimulated by growth factors, hormones, mechanical loading and nutrients, mTORC1 governs energy distribution by turning on/off energy-consuming and energy-generating pathways, depending on cellular energy levels (HOPPE09). This greatly affects ribosome biogenesis, being a process that consumes approximately 80% of cellular energy and materials (VONWALDEN19). Upon stimulation by mTORC1, ribosome biogenesis is initiated with the assembly of the preinitiation complex (PIC) (VONWALDEN19, FIGUEIREDO19). The PIC consists of the RNA polymerase I complex, the SL-1 complex and the TIF-1A/RRN3 protein (FIGUEIREDO19, VONWALDEN19). Further, the upstream binding factor (UBF) and c-Myc have vital roles in the initiation, where UBF binds the PIC to the rDNA promoter and c-Myc as a transcription factor (FIGUEIREDO19, VONWALDEN19). Upon initiation, Pol I transcribes the 47S pre-rRNA, and this is seen as the rate-limiting step (VONWALDEN19, FIGUEIREDO19). The 47S pre-rRNA is processed into the 18S, 5.8S and 28S rRNA (MULLINEUX12, VONWALDEN19), while the last rRNA 5S is transcribed by Pol III, and the ribosomal proteins by Pol II (MULLINEUX12, VONWALDEN19). These events facilitates increased ribosomal content of each skeletal muscle cell, subsequently allowing a higher basal- and roof for MPS, assumed to predict long term adaptations to RT (FIGUEIREDO19/18). Since a blunted ribosome biogenesis has been observed despite increasing mechanical loading (HAMMARSTROM20), targeting hormonal and nutritional signals may ameliorate this. High glucose treatment did promote ribosome biogenesis compared to a repressed ribosome biogenesis with low glucose treatment and glucose starvation, in mice and cell cultures (?)(ZHAI12).

Thus, though it seems apparent that sufficient daily intake of glucose/carbohydrates is necessary for an optimal ribosome biogenesis (ZHAI12, TANAKA18), it is unclear whether there are acute effects on ribosome biogenesis *per se* with ingesting glucose before and after RT. Further, there are equivocal observations on insulin's impact on ribosome biogenesis and MP, where previous meta-analyses suggest insulin to act in concert with amino acid intake to enhance MPS, and insulin may reduce skeletal muscle protein breakdown when amino acids are scarce (ABDULLA16). Therefore, to investigate the relationship between glucose, RT and ribosome biogenesis we recruited 16 young, moderately trained male and female individuals to heavy unilateral resistance training for two weeks. They exercised one leg with glucose and protein, and the other leg with placebo and protein ingestion before and after each RT-session. Our main purpose was to investigate whether there are acute effects of glucose- compared to placebo ingestion before and after RT. To assess this purpose, we measured changes in ribosome biogenesis as accumulation of total RNA, expression of the four mature rRNA's (18S, 5.8S, 28S, 5S), 47S pre-rRNA, and levels of proteins associated with transcription initiation (c-Myc, UBF, rpS6?). As secondary outcomes, we measured the developement in skeletal muscule performance during the intervention, and skeletal muscle recovery after the final RT session, as temporal changes in maximal isometric and isokinetic torque. We hypothesised that RT would indeed lead to a prominent accumulation of total RNA, increased expression of rRNA and associated proteins. Further, we hypothesised that RT with glucose and protein would induce an enhanced response compared to placebo, in markers of ribosome biogenesis, as well as muscular performance and -recovery.

>>>>

Regular resistance exercise training (RT) increases muscle mass and muscle strength, due to the adaptations stimulated by  mechanical strain. Unfortunately, responses to RT are not uniform across the population. Studies have shown RT-induced muscle growth to vary widely between individuals, with as much as 10-15% showing considerably impaired growth (ALVAREZ18;MANN14;THALACKER13). This may be explained by genetics, epigenetics, and/or an unfavourable internal physiological milieu (FIGUEIREDO21;HAMMARSTROM20;THALACKER13). While tweaking training modalties and -protocols induces different responses in different populations (HAMMARSTROM20), little is known regarding optimisation of RT to individual phenotypes. Generally, meta-analyses favour a moderate volume of RT over low volume RT regarding muscle growth (KRIEGER09;RHEA03;SCHOENFELD17), substantiated by recent RCTs (HAMMARSTROM20;SCHOENFELD19). Although Hammarström and colleagues (2020) found that a moderate volume led to more prominent gains in muscle mass, approximately 50% of the participants did not experience true benefits of moderate volume compared to low, with a blunted muscle growth. Due to the variations observed in muscle growth responses despite standardising RT protocols (HAMMARSTROM20;SCHOENFELD17;STEC16), it is apparent that increasing RT volume alone does not convert low-responders to high-responders. Therefore, other means than RT *per se* seems necessary to circumvent this discrepancy.

Indeed, ingesting nutrients such as protein- and creatine-supplements effectively optimise RT (CERMAK12;LANHERS15;LANHERS17). However, it remains unknown if other nutritional adjuvants such as glucose can increase the efficacy of RT. This is surprising, as glucose is the preferred energy substrate of the contracting skeletal muscle during strenuous exercise and a major supplier of energy to cells via ATP synthesis (MUL15;TANAKA18). Energy availability is important to regulate central processes of muscle growth, specifically ribosome biogenesis, a key factor for increased muscle protein synthesis (MPS) (FIGUEIREDO19;HAMMARSTROM20;VONWALDEN19;TANAKA18). Interestingly, variations in RT-induced responses have been linked to ribosome biogenesis, where participants responding poorly to RT also showed a blunted ribosome biogenesis (HAMMARSTROM20;STEC16). A blunted ribosome biogenesis negates responses to RT by failure to produce novel ribosomes, thus an inability to increase the skeletal muscle cell’s translational capacity (VONWALDEN19;FIGUEIREDO19). Therefore, investigating the effects of combined RT and glucose ingestion may provide valuable insight into the regulation of ribosome biogenesis.

Ribosome biogenesis is suggested to be regulated by the mammalian target of rapamycin complex 1 (mTORC1) network (KUSNADI15;VONWALDEn19;TANAKA18). Stimulated by growth factors, hormones, mechanical loading and nutrients, mTORC1 governs energy distribution by turning on/off energy-consuming and energy-generating pathways, depending on cellular energy levels (HOPPE09). This greatly affects ribosome biogenesis, being a process that consumes approximately 80% of cellular energy and materials (VONWALDEN19;TANAKA18). Upon stimulation by mTORC1, assembly of the preinitiation complex (PIC) marks the initiation of ribosome biogenesis (VONWALDEN19;FIGUEIREDO19). The PIC consists of the RNA polymerase I complex, the SL-1 complex and the TIF-1A/RRN3 protein (FIGUEIREDO19, VONWALDEN19). Further, the upstream binding factor (UBF) and c-Myc have vital roles in the initiation, where UBF binds the PIC to the rDNA promoter and c-Myc as a transcription factor (FIGUEIREDO19, VONWALDEN19). Pol I transcribes the 47S pre-rRNA, suggested as the rate-limiting step in ribosome biogenesis (VONWALDEN19, FIGUEIREDO19). The 47S pre-rRNA is processed into the 18S, 5.8S and 28S rRNA (MULLINEUX12, VONWALDEN19), while the last rRNA 5S is transcribed by Pol III, and the ribosomal proteins by Pol II (MULLINEUX12, VONWALDEN19). Importantly, these events facilitates increased ribosomal content of each skeletal muscle cell, subsequently allowing a higher basal- and roof for MPS, suggested to predict long term adaptations to RT (HAMMARSTROM21). Since a blunted ribosome biogenesis has been observed to reduce/inhibit RT responses, despite increasing mechanical loading (HAMMARSTROM20), targeting hormonal and/or nutritional signals may ameliorate this. Previously, high glucose treatment promoted ribosome biogenesis compared to a repressed ribosome biogenesis with low glucose treatment and glucose starvation in mice and cell cultures (MARIAPPAN11;TANAKA15;ZHAI12), making glucose an interesting player in this perspective.

Thus, though it seems apparent that sufficient daily intake of glucose/carbohydrates is necessary for an optimal ribosome biogenesis (TANAKA18), it is unclear whether there are acute effects on ribosome biogenesis *per se* with ingesting glucose before and after RT. Further, there are equivocal observations on insulin's impact on ribosome biogenesis and MP, where previous meta-analyses suggest insulin to act in concert with amino acid intake to enhance MPS, and insulin may reduce skeletal muscle protein breakdown when amino acids are scarce (ABDULLA16). Therefore, to investigate the relationship between glucose, RT and ribosome biogenesis we recruited 16 healthy young, moderately trained male and female individuals to moderate volume unilateral resistance training for two weeks. They exercised one leg with glucose and protein, and the other leg with placebo and protein ingestion before and after each RT-session. Our main purpose was to investigate whether there are acute effects of glucose- compared to placebo ingestion before and after RT. To assess this purpose, we measured changes in ribosome biogenesis as accumulation of total RNA, expression of the four mature rRNA's (18S, 5.8S, 28S, 5S), 47S pre-rRNA, and levels of proteins associated with transcription initiation (c-Myc, UBF, rpS6?). As secondary outcomes, we measured the development in skeletal muscle performance during the intervention, and skeletal muscle recovery after the final RT session, as temporal changes in maximal isometric and isokinetic torque. We hypothesised that RT would indeed lead to a prominent accumulation of total RNA, increased expression of rRNA and associated proteins. Further, we hypothesised that RT with glucose and protein would induce an enhanced response compared to placebo, in markers of ribosome biogenesis, as well as muscular performance and -recovery.
